\bf Model Visualization Techniques for a Social Network Model

### 2018-09-28 paper-child.Rnw
some section 1 changes made. first paragraph(s) totally re-written.

\section{Introduction \& Background}

Social networks, such as collaboration networks between academic researchers or friendship networks have been studied for decades. Statistical models such as exponential random graph models (ERGM) and latent space models (LSM) are often used when studying one observation of a network when trying to model its structure \citep{ergm, lsm}. Social networks can also be observed at many points in time, and ERGM or LSM may not be appropriate for modelling the changes because they do not model network change as time passes. When studying \textit{dynamic networks} observed at many time points, we need a model that allows for variation over time. Models for dynamic social networks hold a great deal of potential because of how realistic they can be: social networks evolve over time as edges are formed and dissolved and new nodes join. Modeling the underlying mechanisms that create network changes in time is very complex but also provides potential to uncover hidden truths.

\par One set of models for dynamic social networks, are stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. \hh{
These models allow us to add node-level covariates into the model by incorporating structural information on a single node, such as in-degree or out-degree, and information derived from pairs of nodes by comparing values of covariates. For example, 
} 
%These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information. Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example 
we expect people with common interests to be more likely to form relationships. SAOMs allow us to incorporate this information on shared interests in the modeling process.

\par Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically. Likelihood functions quickly become very complex due to the dependency structure inherent in the data, and so computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models. SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}. This software is a considerable contribution to the field of social network analysis, but the parameter estimation process is largely hidden from the user. In order to better understand the model-fitting process of \texttt{RSiena}, we visualize the underlying methods and structures, and ultimately aim to help researchers working with SAOMs better understand the structure of and results from these models.

\par To learn more about the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the algorithms fitting the SAOMs. SAOMs are a model family that can reap many benefits from the application of model visualization techniques. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Visualizing the CTMC can provide researchers with more insight into the underlying features of the model. In addition, SAOMs can include a great deal of parameters in the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the corresponding parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model. Furthermore, the estimation of the parameters in a SAOM rely on convergence checks based on simulations from the fitted model. These simulations are hidden and unsaved by default for computational efficiency. With the help of both static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

@@

We provide a brief introduction to the structure of network data and common network visualization methods. \\

\noindent \textbf{Network Data Structure:} Network data across fields have similar data structures. There are always units of observation and connections between those units. In a social network, the units of observations, called \textit{nodes} or \textit{actors}, may be people, while the connections, called \textit{edges} or \textit{ties}, are the relationships between people. Networks often change over time, like when new relationships are formed in a social network. The nodes and edges themselves can also have inherent variables of interest, e.g.\ the age, gender, and political affiliation of people in a social network and the type of relationship (friends, married, etc.). \\

>>>>

\section{Introduction \& Background}
% peices of old intro
%Social networks, such as collaboration networks between academic researchers or friendship networks 
% new intro 
%As humans and researchers, it is our instint to try to learn as much as we can about the world around us. 
As members of the human family, one of the hardest things we try to do is understand our relationships with one another, whether as friends, family, or colleagues. These relationships, or \emph{social networks}, have been studied for decades, in every format from self-help books to publications in academic journals. There are even statistical models, such as exponential random graph models (ERGM) and latent space models (LSM), that strive to capture the underlying relationship-generating mechanism \citep{ergm, lsm}. These static models, however, are not optimal for modelling ``real-world" social networks because they do not account for changes in the network as time passes. \textit{Dynamic social networks} are more realistic because they contain multiple observations of the same network over time, and models for dynamic social networks hold a great deal of promise because of their potential to capture realistic mechanisms. Our social networks evolve over time as new relationships are made, and being able to model these evolutionary mechanisms can uncover hidden truths about our world.

\par One set of models for dynamic social networks, are stochastic actor-oriented models (SAOMs), introduced in \citet{saompaper}. These models incorporate node-level covariate effects in addition to the traditional structural effects, such as in-degree or out-degree. 
%These models are different from other dynamic network models because they allow us to add node level information into the model with the traditional network structure information. Adding node-level information into the model is a more intuitive approach to model changes in a social network because, for example 
With these models, we can examine our assumptions that people with common interests are more likely to become friends. For example, we may assume that high schoolers who drink heavily are more likely to be friends with others who drink heavily. SAOMs allow us to incorporate this behavioral information on shared interests into the modeling process.

\par Unlike other network models, SAOMs are not analytically tractable. Likelihood functions quickly become very complex due to the dependency structure inherent in the data, and so computational methods, such as Markov chain Monte Carlo (MCMC), are used to fit models. %SAOMs are typically fit to dynamic social network data using a series of MCMC phases for finding method of moments (MoM) estimates of model parameters. 
In order to estimate the parameters of a SAOM, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena} by \citet{RSiena}. This software is a considerable contribution to the field of social network analysis, but the extensive parameter estimation process is largely hidden from the user. In order to better understand the fitting process in \texttt{RSiena}, we use model visualization, or model-vis \citep{modelvis}, to examine the underlying methods, hidden structures, and simulation results. Ultimately, we aim to help researchers working with SAOMs better understand the structure, behavior, fit, and results of these models.

In the remainder of this introduction, we first describe the structure of network data and different network visualization techniques. Then, we carefully define the many moving pieces in SAOMs. Finally, we briefly ouline the SAOM fitting process to a set of dynamic network data.

%\par To learn more about the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the algorithms fitting the SAOMs. SAOMs are a model family that can reap many benefits from the application of model visualization techniques. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Visualizing the CTMC can provide researchers with more insight into the underlying features of the model. In addition, SAOMs can include a great deal of parameters in the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the corresponding parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model. Furthermore, the estimation of the parameters in a SAOM rely on convergence checks based on simulations from the fitted model. These simulations are hidden and unsaved by default for computational efficiency. With the help of both static and dynamic visualizations, we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

@@

%We provide a brief introduction to the structure of network data and common network visualization methods. \\

%\noindent \textbf{Network Data Structure:} 
Network data across fields have similar data structures. There are always units of observation and connections between those units. In a social network, the units of observations, called \textit{nodes} or \textit{actors}, are people, while the connections, called \textit{edges} or \textit{ties}, are their relationships. %Networks change over time, like when new relationships are formed in a social network. 
Other data on a network include inherent variables of interest on the nodes and edges themselves, e.g.\ someone's age, gender, and political affiliation, the type of relationship (friendship, coworkers, etc) that an edge represents, in addition to the changing structure of the network in time. \\
